---
title: "Hunger in America is a Policy Crisis"
author: "Mark Hamer"
date: "May 3, 2026"
output:
  pdf_document:
    toc: false
    number_sections: false
    fig_caption: true
  html_document:
    toc: false
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 5.5
)
library(usmap)
library(ggplot2)
library(dplyr)

```

## Introduction

47 million Americans faced food insecurity in 2023. That is 1 in 7 
households. This report is not about hunger in developing countries. 
It is about hunger here. In every state. In working families.

The United Nations tracks food insecurity as a global crisis. The data 
shows it is also a domestic one. The difference is that in America, 
hunger is not caused by drought or war. It is caused by policy. 
Specifically, a nutrition assistance system that was designed with a 
hard cutoff that punishes families for earning more.

This report makes three arguments. First, food insecurity in America 
is concentrated but universal. No state is fully exempt. Second, a large portion of the people 
who go hungry are largely working people who earn too much to qualify 
for help but too little to afford food. Third, the gap between what 
SNAP provides and what adequate nutrition costs is large enough to 
matter and small enough to fix.

The data comes from Feeding America, the USDA, and the U.S. Census 
Bureau. 

## Where Hunger Happens

Food insecurity is not evenly distributed. Some states are significantly 
worse than others. The map below shows food insecurity rates by state 
using 2023 data. The concentration in the South is not random. It 
reflects decades of lower wages, weaker labor protections, and higher 
poverty rates in that region.

```{r}
library(usmap)
library(ggplot2)
library(dplyr)

fi_data <- data.frame(
  state = c(
    "AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "DC",
    "FL", "GA", "HI", "ID", "IL", "IN", "IA", "KS", "KY",
    "LA", "ME", "MD", "MA", "MI", "MN", "MS", "MO", "MT",
    "NE", "NV", "NH", "NJ", "NM", "NY", "NC", "ND", "OH",
    "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT",
    "VT", "VA", "WA", "WV", "WI", "WY"
  ),
  food_insecurity_rate = c(
    17.5, 14.5, 14.4, 19.3, 13.7, 12.7, 14.3, 13.0, 12.4,
    14.4, 14.9, 14.2, 12.7, 13.1, 15.1, 12.0, 14.0, 16.6,
    18.1, 13.8, 13.3, 11.5, 15.4, 10.4, 19.4, 15.4, 12.3,
    14.5, 15.1, 10.7, 11.7, 16.6, 14.5, 15.0,  9.9, 15.3,
    18.4, 14.2, 13.2, 12.3, 14.1, 12.4, 15.6, 17.6, 14.2,
    12.2, 12.1, 13.2, 15.7, 11.8, 15.4
  )
)

plot_usmap(data = fi_data, values = "food_insecurity_rate", color = "white", linewidth = 0.3) +
  scale_fill_gradient(
    low    = "#fcbba1",
    high   = "#67000d",
    name   = "Share of residents\nfacing food insecurity",
    limits = c(0, 20),
    breaks = c(0, 5, 10, 15, 20),
    labels = c("0%", "5%", "10%", "15%", "20%")
  ) +
  labs(
    title    = "Geography Still Determines Who Goes Hungry in America",
    subtitle = "Southern states consistently report food insecurity rates far above the national average.\nU.S. average: 13.5%  |  Worst states: MS 19.4%, AR 19.3%, OK 18.4%, LA 18.1%, TX 17.6%",
    caption  = "Source: Feeding America, Map the Meal Gap 2025 (2023 data)\nNote: No state falls below 9%. Every state on this map has a hunger problem."
  ) +
  theme(
    legend.position  = "right",
    plot.title       = element_text(size = 13, face = "bold", hjust = 0.5),
    plot.subtitle    = element_text(size = 9, color = "gray30", hjust = 0.5),
    plot.caption     = element_text(size = 8, color = "gray50"),
    legend.title     = element_text(size = 9),
    legend.text      = element_text(size = 8),
    panel.background = element_rect(fill = "white", color = NA),
    plot.background  = element_rect(fill = "white", color = NA)
  )
```
Mississippi, Arkansas, Oklahoma, Louisiana, and Alabama all exceed 17%. 
North Dakota is the lowest at 9.9%. Every single state is above 9%. 
There is no state where hunger is not a problem.

---

## Who Hunger Affects

Food insecurity does not affect everyone equally. Some groups face it 
at rates far above the national average. Understanding who is most 
affected is the first step to understanding why the system fails them. 
The chart below breaks down food insecurity rates by age group and 
household type using 2023 national data.

```{r}
library(ggplot2)

demo_data <- data.frame(
  group = c(
    "Women with children",
    "Men with children",
    "Children under 18",
    "Women living alone",
    "Working age adults",
    "Men living alone",
    "Adults 50 to 59",
    "Seniors 60 and older"
  ),
  rate = c(35.3, 23.3, 18.2, 16.8, 15.1, 13.9, 11.8, 8.2),
  highlight = c(TRUE, FALSE, TRUE, FALSE, FALSE, FALSE, FALSE, FALSE)
)

demo_data$group <- factor(
  demo_data$group,
  levels = demo_data$group[order(demo_data$rate)]
)

ggplot(demo_data, aes(x = rate, y = group, fill = highlight)) +
  geom_col(width = 0.6) +
  geom_vline(
    xintercept = 13.5,
    linetype   = "dashed",
    color      = "gray50",
    linewidth  = 0.5
  ) +
  annotate("text",
    x = 13.7, y = 0.5,
    label = "National avg 13.5%",
    size  = 2.6, color = "gray50", hjust = 0
  ) +
  geom_text(
    aes(label = paste0(rate, "%")),
    hjust    = -0.15,
    size     = 3.2,
    fontface = "bold",
    color    = "gray20"
  ) +
  scale_fill_manual(
    values = c("TRUE" = "#67000d", "FALSE" = "#cb4c4c")
  ) +
  scale_x_continuous(
    limits = c(0, 44),
    labels = function(x) paste0(x, "%")
  ) +
  labs(
    title    = "Children and Single Mothers Face the Highest Rates of Food Insecurity",
    subtitle = "Food insecurity falls hardest on children and women raising families alone.\nAll rates are national averages for 2023.",
    x        = "Food insecurity rate (%)",
    y        = NULL,
    caption  = "Sources: Feeding America MMG 2025; USDA ERS Household Food Security Report 2023"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title         = element_text(face = "bold", size = 12),
    plot.subtitle      = element_text(size = 9, color = "gray40", lineheight = 1.4),
    plot.caption       = element_text(size = 8, color = "gray50"),
    legend.position    = "none",
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    panel.grid.major.x = element_line(color = "gray93"),
    plot.margin        = margin(10, 20, 10, 20)
  )
```

Women raising children alone face a 35.3% food insecurity rate. That 
is more than 1 in 3 households. Children under 18 face an 18.2% rate. 
Seniors face the lowest rate at 8.2%, partly because Social Security 
and Medicare provide more stable support than SNAP does for working 
age adults.

The child rate matters most for what comes next. Food insecurity in 
childhood does not stay in childhood. It follows children into 
adulthood through lower educational attainment, worse health outcomes, 
and lower earnings. The chart below shows that progression.

---

A food insecure child is more likely to fall behind in school, less 
likely to graduate, more likely to face chronic illness, and more 
likely to be food insecure as an adult. This is not a temporary 
hardship that resolves itself. It is a cycle that the current system 
reproduces.


```{r}
library(ggplot2)
library(grid)

childhood <- data.frame(
  x     = 0,
  y     = 3,
  label = "Children facing\nfood insecurity\n18.2% nationally",
  type  = "start"
)

outcomes <- data.frame(
  x = 1,
  y = c(5, 4, 3, 2, 1),
  label = c(
    "26% more likely to be\nbelow grade level in reading",
    "22% less likely to\ngraduate high school",
    "35% higher rates of\nchronic illness in adulthood",
    "20% lower lifetime\nearnings on average",
    "28% face food insecurity\nas adults"
  ),
  type = "outcome"
)

# connection lines from start node to each outcome
connections <- data.frame(
  x    = rep(0.05, 5),
  xend = rep(0.85, 5),
  y    = rep(3, 5),
  yend = outcomes$y
)

ggplot() +
  geom_segment(
    data = connections,
    aes(x = x, xend = xend, y = y, yend = yend),
    color     = "#cb4c4c",
    linewidth = 0.7,
    alpha     = 0.5
  ) +
  geom_rect(
    data = childhood,
    aes(xmin = -0.3, xmax = 0.3, ymin = y - 0.55, ymax = y + 0.55),
    fill  = "#67000d",
    color = NA
  ) +
  geom_text(
    data = childhood,
    aes(x = x, y = y, label = label),
    color    = "white",
    size     = 3,
    fontface = "bold",
    lineheight = 1.3
  ) +
  geom_rect(
    data = outcomes,
    aes(xmin = 0.7, xmax = 1.3, ymin = y - 0.38, ymax = y + 0.38),
    fill  = "#cb4c4c",
    color = NA,
    alpha = 0.9
  ) +
  geom_text(
    data = outcomes,
    aes(x = x, y = y, label = label),
    color      = "white",
    size       = 2.8,
    lineheight = 1.3
  ) +
  annotate("text",
    x = 0.5, y = 6,
    label    = "leads to",
    size     = 3.5,
    color    = "gray50",
    fontface = "italic"
  ) +
  scale_x_continuous(limits = c(-0.5, 1.6)) +
  scale_y_continuous(limits = c(0.3, 6.5)) +
  labs(
    title    = "Food Insecure Children Are Less Likely to Become Fully Independent Adults",
    subtitle = "Childhood food insecurity is linked to worse outcomes across education, health, and earnings.\nThese are not isolated effects. They compound over a lifetime.",
    caption  = "Sources: USDA ERS; Urban Institute; Annie E. Casey Foundation; Center on Budget and Policy Priorities\nNote: Outcome statistics reflect elevated rates among populations who experienced childhood food insecurity."
  ) +
  theme_void() +
  theme(
    plot.title    = element_text(face = "bold", size = 12, hjust = 0.5,
                                 margin = margin(b = 6)),
    plot.subtitle = element_text(size = 9, color = "gray40", hjust = 0.5,
                                 lineheight = 1.4, margin = margin(b = 10)),
    plot.caption  = element_text(size = 7, color = "gray50", hjust = 0.5,
                                 lineheight = 1.3, margin = margin(t = 10)),
    plot.margin   = margin(15, 20, 15, 20),
    plot.background = element_rect(fill = "white", color = NA)
  )
```

This brings us to the structural question. If these outcomes are 
predictable and measurable, why does the system continue to produce 
them? The answer is in how SNAP is designed, who it covers, and how 
much it provides. The charts that follow show each of those failures 
in turn.

---

## Why It Happens

The map shows where hunger is worst. This chart shows why. States with 
higher poverty rates consistently have higher food insecurity rates. 
This is not a coincidence. It is a structural relationship. When 
families cannot afford basic expenses, food is often the first thing 
that gets cut. Each point on the chart below is a state. The line 
shows the national trend. 

```{r packages}
library(ggplot2)
library(ggrepel)

state_data <- data.frame(
  state = c(
    "Alabama", "Alaska", "Arizona", "Arkansas", "California",
    "Colorado", "Connecticut", "Delaware", "Florida", "Georgia",
    "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa",
    "Kansas", "Kentucky", "Louisiana", "Maine", "Maryland",
    "Massachusetts", "Michigan", "Minnesota", "Mississippi", "Missouri",
    "Montana", "Nebraska", "Nevada", "New Hampshire", "New Jersey",
    "New Mexico", "New York", "North Carolina", "North Dakota", "Ohio",
    "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "South Carolina",
    "South Dakota", "Tennessee", "Texas", "Utah", "Vermont",
    "Virginia", "Washington", "West Virginia", "Wisconsin", "Wyoming"
  ),
  food_insecurity = c(
    17.5, 14.5, 14.4, 19.3, 13.7, 12.7, 14.3, 13.0, 14.4, 14.9,
    14.2, 12.7, 13.1, 15.1, 12.0, 14.0, 16.6, 18.1, 13.8, 13.3,
    11.5, 15.4, 10.4, 19.4, 15.4, 12.3, 14.5, 15.1, 10.7, 11.7,
    16.6, 14.5, 15.0,  9.9, 15.3, 18.4, 14.2, 13.2, 12.3, 14.1,
    12.4, 15.6, 17.6, 14.2, 12.2, 12.1, 13.2, 15.7, 11.8, 15.4
  ),
  poverty_rate = c(
    16.9, 10.7, 13.5, 16.8, 12.5,  9.1, 10.4,  9.9, 13.5, 13.9,
    10.6, 11.5, 12.0, 12.2,  9.6, 11.0, 16.6, 18.6, 10.9,  8.0,
     9.7, 13.6,  8.9, 18.7, 13.4, 13.5, 10.1, 12.8,  7.0,  9.0,
    19.8, 13.4, 14.2,  9.2, 13.5, 16.3, 12.3, 11.4, 10.4, 15.0,
    11.1, 14.6, 14.7,  8.9,  9.5,  9.5,  9.4, 17.0, 10.5, 10.5
  )
)

highlight_states <- c(
  "Mississippi", "Arkansas", "Louisiana",
  "Oklahoma", "West Virginia", "Minnesota", "New Hampshire"
)

state_data$highlight <- state_data$state %in% highlight_states

ggplot(state_data, aes(x = poverty_rate, y = food_insecurity)) +
  geom_point(
    data = subset(state_data, !highlight),
    color = "gray75", size = 2.5, alpha = 0.6
  ) +
  geom_point(
    data = subset(state_data, highlight),
    color = "#b22222", size = 3.5, alpha = 0.95
  ) +
  geom_smooth(
    method = "lm", se = TRUE,
    color = "gray40", fill = "gray90",
    linewidth = 0.6, alpha = 0.3
  ) +
  geom_vline(
    xintercept = mean(state_data$poverty_rate),
    linetype = "dashed", color = "gray60", linewidth = 0.4
  ) +
  geom_hline(
    yintercept = mean(state_data$food_insecurity),
    linetype = "dashed", color = "gray60", linewidth = 0.4
  ) +
  annotate("text",
    x = mean(state_data$poverty_rate) + 0.2,
    y = 10.6,
    label = "Avg poverty",
    size = 2.5, color = "gray50", hjust = 0
  ) +
  annotate("text",
    x = 7.2,
    y = mean(state_data$food_insecurity) + 0.3,
    label = "Avg food insecurity",
    size = 2.5, color = "gray50", hjust = 0
  ) +
  geom_label_repel(
    data = subset(state_data, highlight),
    aes(label = state),
    size = 2.8,
    color = "#b22222",
    fill = "white",
    label.size = 0.2,
    min.segment.length = 0,
    box.padding = 0.5,
    show.legend = FALSE
  ) +
  scale_x_continuous(
    labels = function(x) paste0(x, "%"),
    limits = c(6.5, 21)
  ) +
  scale_y_continuous(
    labels = function(x) paste0(x, "%"),
    limits = c(9.5, 20.5)
  ) +
  labs(
    title    = "States With Higher Poverty Rates Also Face Higher Food Insecurity",
    subtitle = "States with higher poverty rates consistently experience higher food insecurity.\nEach point represents a state. Highlighted states are the clearest outliers.",
    x        = "Poverty rate (%)",
    y        = "Food insecurity rate (%)",
    caption  = "Sources: Feeding America MMG 2025; U.S. Census Bureau ACS 2023"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold", size = 12),
    plot.subtitle    = element_text(size = 9, color = "gray40"),
    plot.caption     = element_text(size = 8, color = "gray50"),
    legend.position  = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray93")
  )
```

New Mexico has a 19.8% poverty rate and a 16.6% food insecurity rate. 
New Hampshire has a 7% poverty rate and a 10.7% food insecurity rate. 
The relationship holds across every region. Poverty and hunger move 
together. Any serious attempt to reduce food insecurity has to address 
the economic conditions that create it.

---


## What Adequate Nutrition Actually Costs

The USDA publishes a monthly cost estimate for the cheapest diet that 
still meets nutritional guidelines. It is called the Thrifty Food Plan. 
It assumes all meals are cooked at home and represents the absolute 
floor of adequate nutrition. The chart below compares what that costs 
for four common household types against what the average SNAP benefit 
actually provides.

```{r}
library(ggplot2)
library(scales)

household_nutrition <- data.frame(
  household = c(
    "Single adult",
    "Single adult\nwith one child",
    "Two adults",
    "Two adults\nwith two children"
  ),
  tfp_cost  = c(330, 490, 605, 984),
  avg_snap  = c(183, 366, 366, 732)
)

household_nutrition$gap <- household_nutrition$tfp_cost - household_nutrition$avg_snap

household_nutrition$household <- factor(
  household_nutrition$household,
  levels = c(
    "Single adult",
    "Single adult\nwith one child",
    "Two adults",
    "Two adults\nwith two children"
  )
)

# reshape to long for grouped bars
nutrition_long <- rbind(
  data.frame(
    household = household_nutrition$household,
    amount    = household_nutrition$avg_snap,
    type      = "Average SNAP benefit"
  ),
  data.frame(
    household = household_nutrition$household,
    amount    = household_nutrition$tfp_cost,
    type      = "Minimum nutritious diet cost"
  )
)

nutrition_long$type <- factor(
  nutrition_long$type,
  levels = c("Average SNAP benefit", "Minimum nutritious diet cost")
)

# gap labels for annotation
gap_labels <- data.frame(
  household = household_nutrition$household,
  tfp_cost  = household_nutrition$tfp_cost,
  gap       = household_nutrition$gap,
  label     = paste0("$", household_nutrition$gap, " gap")
)

ggplot(nutrition_long, aes(x = household, y = amount, fill = type)) +
  geom_col(position = "dodge", width = 0.6) +
  geom_text(
    data = gap_labels,
    aes(
      x     = household,
      y     = tfp_cost + 25,
      label = label,
      fill  = NULL
    ),
    size     = 3,
    fontface = "bold",
    color    = "#8b0000"
  ) +
  scale_fill_manual(
    values = c(
      "Average SNAP benefit"         = "#4a9e6b",
      "Minimum nutritious diet cost" = "#c0392b"
    ),
    name = NULL
  ) +
  scale_y_continuous(
    labels = dollar_format(),
    limits = c(0, 1100),
    breaks = seq(0, 1000, 200)
  ) +
  labs(
    title    = "SNAP Benefits Do Not Cover the Cost of a Minimum Nutritious Diet",
    subtitle = "Average SNAP benefits fall short for every household type.\nThe gap ranges from $124 to $252 per month.",
    x        = NULL,
    y        = "Monthly cost ($)",
    caption  = "Sources: USDA Thrifty Food Plan November 2024; USDA FNS SNAP Data 2023\nNote: Thrifty Food Plan costs include household size adjustment factors. SNAP benefit is the average per person ($183) multiplied by household size."
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold", size = 12),
    plot.subtitle    = element_text(size = 9, color = "gray40", lineheight = 1.4),
    plot.caption     = element_text(size = 7, color = "gray50", lineheight = 1.3),
    legend.position  = "bottom",
    legend.text      = element_text(size = 10),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(color = "gray93"),
    axis.text.x      = element_text(size = 9, lineheight = 1.4),
    plot.margin      = margin(15, 20, 15, 20)
  )
```

The average SNAP benefit falls short for every household type. A single 
adult receives an average of $183 per month but needs $330 to meet the 
minimum nutrition standard. A family of four receives $732 on average 
but needs $984. The gap ranges from $124 to $252 per month depending 
on household size. These are not large numbers in the context of a 
federal budget. They are very large numbers for a family trying to eat.

---

## Who Actually Goes Hungry

Food insecurity is often assumed to affect people who do not work. The 
data does not support that assumption. The chart below shows food 
insecurity rates by income level. People below the poverty line face 
the highest rates. But people above the poverty line who do not qualify 
for SNAP face a 28% food insecurity rate. That is nearly double the 
national average. They earn too much for help. They do not earn enough 
to eat.

```{r}
income_data <- data.frame(
  category = c(
    "Below poverty line",
    "Above poverty line,\nbut limited SNAP access (130-185% FPL)",
    "Above 185% FPL,\nno SNAP access"
  ),
  rate = c(39.4, 28.1, 6.2)
)

income_data$category <- factor(
  income_data$category,
  levels = income_data$category[order(income_data$rate)]
)

income_data$highlight <- income_data$category == "Above poverty line,\nbut limited SNAP access (130-185% FPL)"

ggplot(income_data, aes(x = rate, y = category, fill = highlight)) +
  geom_col(width = 0.5) +
  geom_vline(
    xintercept = 13.5,
    linetype = "dashed",
    color = "gray50",
    linewidth = 0.5
  ) +
  annotate("text",
  x = 13.7, y = 3.4,
  label = "National avg 13.5%",
  size = 2.6, color = "gray50", hjust = 0
)+
  geom_text(
    aes(label = paste0(rate, "%")),
    hjust = -0.15, size = 3.2,
    fontface = "bold", color = "gray20"
  ) +
  scale_fill_manual(
    values = c("TRUE" = "#8b4513", "FALSE" = "#cb4c4c")
  ) +
  scale_x_continuous(
    limits = c(0, 46),
    labels = function(x) paste0(x, "%")
  ) +
  labs(
    title    = "Earning More Than the Poverty Line Does Not Mean You Can Afford Food",
    subtitle = "People above the poverty line but below 185% FPL face a 28% food insecurity rate.\nThey earn too much for SNAP. They do not earn enough to eat.",
    x        = "Food insecurity rate (%)",
    y        = NULL,
    caption  = "Source: USDA ERS Household Food Security Report 2023"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title         = element_text(face = "bold", size = 12),
    plot.subtitle      = element_text(size = 9, color = "gray40"),
    plot.caption       = element_text(size = 8, color = "gray50"),
    legend.position    = "none",
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    panel.grid.major.x = element_line(color = "gray93")
  )
```

This group is sometimes called the working poor. They are employed. 
They pay taxes. They do not qualify for SNAP because their income 
exceeds 130% of the federal poverty line. In 2025 that cutoff is 
$20,345 for a single person. A person earning $21,000 per year gets 
no SNAP assistance. They also cannot afford adequate nutrition on that 
income in most parts of the country.

---

## The System Punishes Earning More

SNAP eligibility cuts off at 130% of the federal poverty line. There 
is no gradual reduction. One dollar over the limit and the entire 
benefit disappears. A $1,000 annual raise adds $83 per month to a 
household's income. But crossing the SNAP threshold means losing 
hundreds of dollars in monthly benefits at once. The net result is 
negative. The charts below show the math for every household size.


```{r}

library(ggplot2)
library(scales)
library(dplyr)

waterfall_data <- data.frame(
  household = rep(c("1-Person", "2-Person", "3-Person", "4-Person"), each = 3),
  label = rep(c("Pay raise", "SNAP lost", "Net change"), 4),
  value = c(
     83, -167, -84,
     83, -153, -70,
     83,  -96, -13,
     83,  -70, -17
  ),
  type = rep(c("gain", "loss", "net"), 4)
)

waterfall_data$household <- factor(
  waterfall_data$household,
  levels = c("1-Person", "2-Person", "3-Person", "4-Person")
)

waterfall_data$label <- factor(
  waterfall_data$label,
  levels = c("Pay raise", "SNAP lost", "Net change")
)

waterfall_data$type <- factor(
  waterfall_data$type,
  levels = c("gain", "loss", "net"),
  labels = c("Pay raise", "SNAP benefits lost", "Net change")
)

waterfall_data$fill <- case_when(
  waterfall_data$type == "Pay raise"          ~ "#4a9e6b",
  waterfall_data$type == "SNAP benefits lost" ~ "#c0392b",
  waterfall_data$type == "Net change"         ~ "#2c2c2c"
)

waterfall_data$value_label <- case_when(
  waterfall_data$type == "Pay raise"          ~ paste0("+$", waterfall_data$value, "/mo"),
  waterfall_data$type == "Net change"         ~ paste0("Net: -$", abs(waterfall_data$value), "/mo"),
  TRUE                                        ~ paste0("-$", abs(waterfall_data$value), "/mo")
)

waterfall_data$vjust_val    <- ifelse(waterfall_data$value >= 0, -0.6, 1.5)
waterfall_data$fontface_val <- ifelse(waterfall_data$type == "Net change", "bold", "plain")

ggplot(waterfall_data, aes(x = label, y = value, fill = type)) +
  geom_col(width = 0.45) +
  geom_hline(yintercept = 0, color = "gray40", linewidth = 0.5) +
  geom_text(
    aes(
      label    = value_label,
      vjust    = vjust_val,
      fontface = fontface_val
    ),
    size  = 3,
    color = "gray20"
  ) +
  scale_fill_manual(
    values = c(
      "Pay raise"          = "#4a9e6b",
      "SNAP benefits lost" = "#c0392b",
      "Net change"         = "#2c2c2c"
    ),
    name = NULL
  ) +
  scale_y_continuous(
    labels = dollar_format(),
    limits = c(-240, 160),
    breaks = seq(-200, 100, 50)
  ) +
  facet_wrap(~household, ncol = 4) +
  labs(
    title    = "SNAP Benefit Cliffs Can Offset the Gains From a Raise",
    subtitle = "A $1,000 annual raise adds about $83 per month.\nCrossing the SNAP cutoff can reduce benefits by even more, leaving families with less than before.",
    x        = NULL,
    y        = "Monthly change in resources ($)",
    caption  = "Sources: USDA SNAP max allotments 2025; HHS Federal Poverty Guidelines 2025"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title         = element_text(face = "bold", size = 13, margin = margin(b = 6)),
    plot.subtitle      = element_text(size = 10, color = "gray40",
                                      lineheight = 1.6, margin = margin(b = 16)),
    plot.caption       = element_text(size = 8, color = "gray50", margin = margin(t = 12)),
    strip.text         = element_text(face = "bold", size = 11, margin = margin(b = 10)),
    legend.position    = "bottom",
    legend.text        = element_text(size = 10),
    legend.spacing.x   = unit(0.5, "cm"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_line(color = "gray93"),
    panel.spacing      = unit(2.5, "lines"),
    plot.margin        = margin(15, 25, 15, 25),
    axis.text.x        = element_blank(),
    axis.text.y        = element_text(size = 9),
    axis.title.y       = element_text(size = 10, margin = margin(r = 10))
  )
```

A single person who crosses the cutoff loses $167 per month in net 
resources despite earning more. A two-person household loses $70. 
This is not a design flaw. It is how the program was built. The 
rational response for many families is to avoid the raise entirely. 
The policy does not reward work. It penalizes crossing an arbitrary 
line.

---

## What a Better System Could Look Like

The cliff effect is not inevitable. It is a product of a binary 
eligibility threshold. Benefits either exist at full value or they 
disappear. A gradual phase-out model would reduce benefits slowly as 
income rises rather than eliminating them at a single cutoff point. 
The chart below shows the same $1,000 raise scenario under a gradual 
phase-out model for each household size.


```{r}
library(ggplot2)
library(scales)
library(dplyr)

waterfall_data <- data.frame(
  household = rep(c("1-Person", "2-Person", "3-Person", "4-Person"), each = 3),
  label     = rep(c("Pay raise", "SNAP reduction", "Net change"), 4),
  value     = c(
     83, -40,  43,
     83, -40,  43,
     83, -25,  58,
     83, -20,  63
  ),
  type = rep(c("gain", "loss", "net"), 4)
)

waterfall_data$household <- factor(
  waterfall_data$household,
  levels = c("1-Person", "2-Person", "3-Person", "4-Person")
)

waterfall_data$label <- factor(
  waterfall_data$label,
  levels = c("Pay raise", "SNAP reduction", "Net change")
)

waterfall_data$value_label <- case_when(
  waterfall_data$type == "gain" ~ paste0("+$", waterfall_data$value, "/mo"),
  waterfall_data$type == "net"  ~ paste0("Net: +$", waterfall_data$value, "/mo"),
  TRUE                          ~ paste0("-$", abs(waterfall_data$value), "/mo")
)

waterfall_data$vjust_val    <- ifelse(waterfall_data$value >= 0, -0.6, 1.5)
waterfall_data$fontface_val <- ifelse(waterfall_data$type == "net", "bold", "plain")

ggplot(waterfall_data, aes(x = label, y = value, fill = type)) +
  geom_col(width = 0.45) +
  geom_hline(yintercept = 0, color = "gray40", linewidth = 0.5) +
  geom_text(
    aes(
      label    = value_label,
      vjust    = vjust_val,
      fontface = fontface_val
    ),
    size  = 3.2,
    color = "gray20"
  ) +
  scale_fill_manual(
    values = c(
      "gain" = "#4a9e6b",
      "loss" = "#c0392b",
      "net"  = "#2e7d32"
    ),
    labels = c(
      "gain" = "Pay raise",
      "loss" = "SNAP reduction",
      "net"  = "Net change"
    ),
    name = NULL
  ) +
  scale_y_continuous(
    labels = dollar_format(),
    limits = c(-80, 130),
    breaks = seq(-80, 130, 40)
  ) +
  facet_wrap(~household, ncol = 4) +
  labs(
    title    = "A Gradual Phase-Out Means Every Household Keeps More From a Raise",
    subtitle = "Under a gradual phase-out, SNAP benefits reduce slowly as income rises.\nA $1,000 annual raise produces a net gain for every household size.",
    x        = NULL,
    y        = "Monthly change in resources ($)",
    caption  = "Sources: USDA SNAP 2025; HHS Federal Poverty Guidelines 2025\nNote: SNAP reductions are illustrative based on a proportional taper model."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title         = element_text(face = "bold", size = 13, margin = margin(b = 6)),
    plot.subtitle      = element_text(size = 10, color = "gray40",
                                      lineheight = 1.5, margin = margin(b = 16)),
    plot.caption       = element_text(size = 7, color = "gray50",
                                      lineheight = 1.3, margin = margin(t = 12)),
    strip.text         = element_text(face = "bold", size = 11, margin = margin(b = 10)),
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_line(color = "gray93"),
    panel.spacing      = unit(2.5, "lines"),
    plot.margin        = margin(15, 25, 15, 25),
    axis.text.x        = element_blank(),
    axis.text.y        = element_text(size = 9),
    axis.title.y       = element_text(size = 10, margin = margin(r = 10)),
    legend.position    = "bottom",
    legend.text        = element_text(size = 10)
  )
```

Under a gradual phase-out, a $1,000 raise produces a net gain for 
every household size. The benefit reduction is smaller and the family 
keeps more of what they earn. This is not a radical proposal. It is 
how most phase-out programs in other countries are structured. It 
preserves work incentives without eliminating support for families 
who still need it.

---

## What Reform Would Look Like at the National Level

The opening map showed food insecurity as it exists today. The maps 
below show today alongside a projected outcome if SNAP benefits were 
expanded to meet the Thrifty Food Plan minimum cost. The projection 
applies a conservative 20% reduction in food insecurity rates based 
on research linking adequate benefit levels to improved outcomes. 
Both maps use the same color scale.

```{r}
library(usmap)
library(ggplot2)
library(patchwork)

fi_current <- data.frame(
  state = c(
    "AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "DC",
    "FL", "GA", "HI", "ID", "IL", "IN", "IA", "KS", "KY",
    "LA", "ME", "MD", "MA", "MI", "MN", "MS", "MO", "MT",
    "NE", "NV", "NH", "NJ", "NM", "NY", "NC", "ND", "OH",
    "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT",
    "VT", "VA", "WA", "WV", "WI", "WY"
  ),
  food_insecurity_rate = c(
    17.5, 14.5, 14.4, 19.3, 13.7, 12.7, 14.3, 13.0, 12.4,
    14.4, 14.9, 14.2, 12.7, 13.1, 15.1, 12.0, 14.0, 16.6,
    18.1, 13.8, 13.3, 11.5, 15.4, 10.4, 19.4, 15.4, 12.3,
    14.5, 15.1, 10.7, 11.7, 16.6, 14.5, 15.0,  9.9, 15.3,
    18.4, 14.2, 13.2, 12.3, 14.1, 12.4, 15.6, 17.6, 14.2,
    12.2, 12.1, 13.2, 15.7, 11.8, 15.4
  )
)

fi_projected <- data.frame(
  state = c(
    "AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "DC",
    "FL", "GA", "HI", "ID", "IL", "IN", "IA", "KS", "KY",
    "LA", "ME", "MD", "MA", "MI", "MN", "MS", "MO", "MT",
    "NE", "NV", "NH", "NJ", "NM", "NY", "NC", "ND", "OH",
    "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT",
    "VT", "VA", "WA", "WV", "WI", "WY"
  ),
  food_insecurity_rate = c(
    14.0, 11.6, 11.5, 15.4, 11.0, 10.2, 11.4, 10.4,  9.9,
    11.5, 11.9, 11.4, 10.2, 10.5, 12.1,  9.6, 11.2, 13.3,
    14.5, 11.0, 10.6,  9.2, 12.3,  8.3, 15.5, 12.3,  9.8,
    11.6, 12.1,  8.6,  9.4, 13.3, 11.6, 12.0,  7.9, 12.2,
    14.7, 11.4, 10.6,  9.8, 11.3,  9.9, 12.5, 14.1, 11.4,
     9.8,  9.7, 10.6, 12.6,  9.4, 12.3
  )
)

shared_scale <- scale_fill_gradient(
  low    = "#fcbba1",
  high   = "#67000d",
  name   = "Food insecurity\nrate (%)",
  limits = c(0, 20),
  breaks = c(0, 5, 10, 15, 20),
  labels = c("0%", "5%", "10%", "15%", "20%")
)

map_before <- plot_usmap(
  data      = fi_current,
  values    = "food_insecurity_rate",
  color     = "white",
  linewidth = 0.3
) +
  shared_scale +
  labs(title = "Today") +
  theme(
    legend.position  = "none",
    plot.title       = element_text(size = 13, face = "bold", hjust = 0.5),
    panel.background = element_rect(fill = "white", color = NA),
    plot.background  = element_rect(fill = "white", color = NA)
  )

map_after <- plot_usmap(
  data      = fi_projected,
  values    = "food_insecurity_rate",
  color     = "white",
  linewidth = 0.3
) +
  shared_scale +
  labs(title = "With Expanded SNAP Benefits") +
  theme(
    legend.position  = "right",
    plot.title       = element_text(size = 13, face = "bold", hjust = 0.5),
    legend.title     = element_text(size = 9),
    legend.text      = element_text(size = 8),
    panel.background = element_rect(fill = "white", color = NA),
    plot.background  = element_rect(fill = "white", color = NA)
  )

map_before + map_after +
  plot_annotation(
    title    = "Expanding SNAP to Meet Nutrition Costs Would Visibly Reduce Hunger Nationwide",
    subtitle = "National average falls from 13.5% to an estimated 10.8%. The South sees the largest reductions.\nBoth maps use the same color scale for direct comparison.",
    caption  = "Source: Feeding America MMG 2025 (2023 data); USDA Thrifty Food Plan November 2024\nNote: Projected rates apply a conservative 20% reduction. This is a directional estimate, not a policy simulation.",
    theme = theme(
      plot.title    = element_text(face = "bold", size = 13, hjust = 0.5),
      plot.subtitle = element_text(size = 9, color = "gray30",
                                   hjust = 0.5, lineheight = 1.4),
      plot.caption  = element_text(size = 7, color = "gray50", lineheight = 1.3)
    )
  )
```

---

## Conclusion

Hunger in America is not a scarcity problem. The food exists. The 
system to distribute it exists. What does not work is the policy 
that determines who gets help and how much.

Three changes would make a measurable difference. First, increase 
SNAP benefits to meet the Thrifty Food Plan minimum cost. The gap 
is between $124 and $252 per month depending on household size. 
Second, replace the hard eligibility cutoff with a gradual phase-out 
that reduces benefits slowly as income rises. Third, index benefit 
levels to regional food costs so families in high cost states are 
not penalized for where they live.

None of these changes require a new program. They require adjustments 
to one that already exists. The national food budget shortfall, the 
total additional dollars food insecure Americans need to meet minimum 
nutrition costs, is $22.6 billion per year according to Feeding 
America. That is less than 0.3% of the federal budget.

---




